Conversas Hiperligadas: Novo Paradigma de Comunicação e Colaboração, potenciado pela Tecnologia WebRTC

### 2016-05-10 tese2/14_implementation_signaling.tex
Reviewing implementation.

\section{Signaling Protocol}

The signaling protocol plays a fundamental role on our solution, after the web server validates the user access,
it allows the users to directly connect to \ac{KMS} which is placed in a private network and let the server and users negotiate media types and encoding information to use during the conversation.

With respect to the signaling protocol, it could be implemented using \ac{HTTP} messages but they would transport extra information such as \ac{HTTP} headers and would follow a request-response signaling mechanism which would not be the best option, as multiple ice candidates can arrive at any time. Using a \ac{REST} \ac{API} over \ac{HTTP} for signaling and allowing other methods to be called at the same time would lead to open multiple \ac{TCP} connections at the same time.

Instead if we create a \emph{WebSocket} \ac{API} we only need one \ac{TCP} connection and at the same provide bi-directional communications without additional headers. Moreover, using \emph{WebSocket} allows the server to send messages to the client without the client having to request it, which can be slightly faster than using \ac{HTTP}.

Our signaling protocol consists in send and receiving \ac{JSON} formated messages like listing \ref{lst:msgtype} over WebSockets by both the server and the client.

@@

When a user enters in a group conference, after the page is completely loaded it is created a WebSocket that maintains a connection with our web servers. 
Before creating the web socket, we must identify the user and check if he has permissions to participate in the conference. The user identification is done by retrieving the session id from the cookie provided by the user through the \ac{HTTP} headers then we retrieve all the information need from database in order to check if the user has permissions to join the conference room. It's important to save the user identification before the socket is created because after the the handshake performed by the \emph{WebSocket} protocol\cite{rfc6455} the \ac{HTTP} context is lost.

When the connection is established between the server and the client, a \emph{PeerConnection} is created on the client and immediately after an \ac{WebRTC} endpoint on the server specifying a possible set of \ac{ICE} servers to connect.

The user is asked if he wants to share its camera and microphone, share screen or just receive streams from the server. If the user decides to perform a screen share, \emph{adapter.js}\footnote{\url{https://github.com/Temasys/AdapterJS}(accessed March 15, 2015).} may ask to install a plug-in if the browser doesn't support screen sharing.

If the user decides to share either from camera of screen, \emph{getUserMedia} is called with the correspondent constraints in order to obtain a local stream. Use the constraints presented on listing \ref{lst:constraints}.

@@

A pop-up is raised in order to ask the user to give permission to share those resources. If the resources are shared successfully the user creates an offer like listing \ref{lst:sig01}, sets a \emph{local session description} to its \emph{PeerConnection} and sends it through WebSockets. If the resources are not shared or the user specified to receive stream only, an offer is created specifying the constraints for receiving only video and audio.

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Offer created by client},label={lst:sig01},language=json]
{
	"cmd":"offer",
	"data":{
		"type":"offer",
		"sdp":<sdp>	// this is huge
	}
}
\end{lstlisting}
\end{minipage}

The \emph{local session description} contains the session identifier, codecs, containers, transport protocols and ports used per media type. The \emph{local session description} is useful to conclude if the client is receiving only, which means that \ac{KMS} doesn't need to mix, record and analyze streams coming from the user \ac{WebRTC} endpoint.

The server receive and processes the offer and sets the \emph{remote session description} to its client associated \ac{WebRTC} endpoint. Then a \emph{local session description} like listing \ref{lst:sig02} is created on the server and sent back to the client, after that the server tries to gather \ac{ICE} candidates.

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Answer created by KMS},label={lst:sig02},language=json]
{
	"cmd":"answer",
	"data":{
		"type":"answer",
		"sdp":<sdp>	// this is huge
	}
}
\end{lstlisting}
\end{minipage}

The client receives the server answer, sets the \emph{remote session description} and gets the ice candidates from the \ac{ICE} server.

Subsequently, after a while both the server and client receive the \ac{ICE} candidates that allow the client to connect directly to \ac{KMS} and vice-versa. The candidates are received at the client which sets them to its \emph{PeerConnection}. The same is done on the server which receives the \ac{ICE} candidates like on listing \ref{lst:sig03} from the client and propagates them to \ac{KMS}.

@@

An \ac{ICE} candidate contains an \ac{IP}, port, used transport protocol and an attribute named \emph{sdpMLineIndex} that is used for mapping to the \emph{remote session description} media type.

In addition both intervenients test the connectivity of each \ac{ICE} candidate. When a connection is established the user and server start to interchange stream data but other \ac{ICE} candidates may arrive with better connections, when that happens the connection changes seamlessly.

@@

>>>>

\section{Signaling Protocol}
%RP devias começar com uma introdução a explicar que tiveste de desenvolver um protocolo para coordenar a interacção dos clientes com o servidor e entre clientes (?).
%RP explicar quando é usado (começar chamada, terminar, juntar, chat(?), edição colaborativa (?), etc).

The signaling protocol plays a fundamental role on our solution.
After the web server validates the user access, it allows the users to directly connect to the \ac{KMS}, which is placed in a private network, and lets the server and users negotiate media types and encoding information to use during the conversation.
%RP o it em ``it allows'' é quem? webserver ou signaling protocol?

We considered implementing the signaling protocol using \ac{HTTP} messages.
However, they would transport extra information such as \ac{HTTP} headers and would follow a request-response signaling mechanism which would not be the best option, as multiple ICE candidates can arrive at any time. Using a \ac{REST} \ac{API} over \ac{HTTP} for signaling and allowing other methods to be called at the same time would lead to opening multiple \ac{TCP} connections at the same time.

Instead, if we create a \emph{WebSocket} \ac{API}, we only need one \ac{TCP} connection and, at the same, provide bi-directional communications without additional headers. Moreover, using \emph{WebSocket} allows the server to send messages to the client without the client having to request it, which can be slightly faster than using \ac{HTTP}.
%RP em vez de slighhtly faster não queres antes enfatizar que pode ser antes o servidor a iniciar a comunicação, de forma assincrona?

Our signaling protocol consists of sending and receiving \ac{JSON} formated messages, e.g. listing \ref{lst:msgtype}, over WebSockets by both the server and the client.

@@

When a user enters a group conference, after the page is completely loaded, a WebSocket is created to maintain a connection with our web servers. 
But before creating the web socket, we must identify the user and check if he has permissions to participate in the conference. The user identification is done by retrieving the session id from the cookie provided by the user through the \ac{HTTP} headers.
%RP não é bem o user que fornece o cookie. é o user-agent (browser). Convém ser preciso.
The Web server retrieves all the information needed from the database in order to check if the user has permissions to join that conference room. It is important to save the user identification before the socket is created because, after the handshake is performed by the \emph{WebSocket} protocol\cite{rfc6455}, the \ac{HTTP} context is lost.
%RP socket -> websocket?

When the connection is established between the server and the client, a \emph{PeerConnection} is created on the client and immediately after, an \ac{WebRTC} endpoint is created on the server, specifying a possible set of \ac{ICE} servers to connect.
%RP Usa Application server e não apenas server. Há muitos servers.

At this stage, the Web Client's user is asked if he wants to share its camera and microphone, share screen or just receive streams from the server. If the user decides to perform a screen share, \emph{adapter.js}\footnote{\url{https://github.com/Temasys/AdapterJS}(accessed March 15, 2015).} may ask to install a plug-in if the browser doesn't support screen sharing.
%RP web app user? OU apenas client user?
%RP foot note com exemplos de browsers que aceitem screen sharing?
If the user decides to share either from camera or screen, \emph{getUserMedia} is called with the correspondent constraints in order to obtain a local stream. We use the constraints presented on listing \ref{lst:constraints}.

@@

If the user decides to share his camera or screen, a pop-up is raised in order to ask the user to give permission to share those resources. If the resources are shared successfully, the user creates an offer like listing \ref{lst:sig01}, sets a \emph{local session description} to its \emph{PeerConnection} and sends it through the WebSocket to the Application Server.
%RP usas ``user'' duas vezes, uma para te referires ao utilizador e outra ao browser (user agent).
If the resources are not shared or the user specified to receive stream only, an offer is created specifying the constraints for receiving only video and audio.
%RP offer -> SDP offer?

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Offer created by client},label={lst:sig01},language=json]
{
	"cmd":"offer",
	"data":{
		"type":"offer",
		"sdp":<sdp>	// omitted for brevity
	}
}
\end{lstlisting}
\end{minipage}

The \emph{local session description} contains the session identifier, codecs, containers, transport protocols and ports used per media type. The \emph{local session description} is useful to conclude if the client is receiving only, which means that \ac{KMS} doesn't need to mix, record nor analyze streams coming from the user's \ac{WebRTC} endpoint.

The server receives and processes the offer and sets the \emph{remote session description} to its client associated \ac{WebRTC} endpoint. Then a \emph{local session description}, like the one presented in Listing \ref{lst:sig02}, is created on the server and sent back to the client, after that the server tries to gather \ac{ICE} candidates.
%RP listing -> Listing

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Answer created by KMS},label={lst:sig02},language=json]
{
	"cmd":"answer",
	"data":{
		"type":"answer",
		"sdp":<sdp>	// omitted for brevity
	}
}
\end{lstlisting}
\end{minipage}

The client receives the server answer, sets the \emph{remote session description} and gets the ice candidates from the \ac{ICE} server.
%RP tens ``ice'', ``\ac{ICE}''. Acho que também já vi ``ICE''. Convém ser sempre igual!

Subsequently, after a while both the server and client receive the \ac{ICE} candidates that allow the client to connect directly to \ac{KMS} and vice-versa. The candidates are received at the client which sets them to its \emph{PeerConnection}. The same is done on the server which receives the \ac{ICE} candidates like on listing \ref{lst:sig03} from the client and propagates them to \ac{KMS}.
%RP A figura 4.1 precisa de vir antes deste texto todo, para ser mais fácil de perceber.
%RP também tens de explicar porque razão escolheste colocar o KMS numa rede privada. Se está numa rede privada, usas NAT ou um proxy (socket) para estabelecer a ligação?

@@

An \ac{ICE} candidate contains an \ac{IP}, port, used transport protocol and an attribute named \emph{sdpMLineIndex} that is used for mapping to the \emph{remote session description} media type.
In addition, both intervenients test the connectivity of each \ac{ICE} candidate. When a connection is established, the user and server start to interchange stream data but other \ac{ICE} candidates may arrive with better connections. When that happens, the connection changes seamlessly.

@@

%RP faz sentido haver dois ICE? Se queres colocar 2 dá-lhes nomes diferentes (nem que seja cliente e servidor ou 1 e 2).

%RP não falas nada sobre chats, etc! Não há signalling para isso?

### 2016-05-10 tese2/15_implementation_recording.tex
Reviewing implementation.

\section{Stream Recording}
	In this section we describe our approaches for the stream recording implementation, namely recording on the client side and server side.

\subsection{Client side recording}
	In order to record used shared streams, our first approach consisted on recording video and audio streams into blobs of limited duration. Because each user was recording directly from their shared local resources, we achieved the best stream quality.

	After recording, each block was uploaded to our server through \ac{HTTP} and saved into the file system, the block meta-data was created and inserted on our database. Each block meta-data contained the file's location for the recorded block, the starting date, duration, user identification and group identification. When the file was completely saved the meta-data was advertised to the remaining users. This meta-data was simply used to refresh the user interface, the meta-data was completely discard after that because an huge amount of small blocks meta-data would use more and more memory as time went by.

	The process to play a video was quite simple, the user specified which user and date was intended to play, the server calculated the intersection between the requested date and the block bounds and returned the file to the client.

	Although this idea was fairly simple, we couldn't achieve seamless sequential block switching. Downloading the video file always toke a noticeable amount of time. To solve this problem while we were playing a block, the next one was downloading in parallel so when the current block finished playing we would have an available block to play.

	Block switching became more acceptable, but switching the \ac{URL} always produced a flash. We solved this problem by having two layers and set the \ac{URL} to the back layer, the front layer frozen with the last frame, then we changed the back layer to front.

	After we implemented our recording solution using this approach, we tested locally and remotely. For remote connections we observed a fairly high bandwidth usage mainly because blocks were both sent and received at maximum quality.

\subsection{Server side recording to file system}

	Before recording any type of stream we had to analyze the user media offer in order to check if a video as really being received by \ac{KMS}, otherwise if we would not verify the user's offer, the recording video would be black.

	The streaming content received on \ac{KMS} was already compressed due to \ac{WebRTC}'s exchanged quality of service metrics data. As a direct consequence our recorder solution used on most cases less disk space per block, but would never use more storage than client side recording. \ac{KMS} allows recording files using \emph{webm} and \emph{mp4} containers.

	With server side recording, the user would maintain always the same stream \ac{URL} even if it is playing real-time video or reproducing recorded video. When a user desires to play recorded video, a \emph{webSocket} message is sent specifying the time and the intended user id, the group identification is not sent because it is already associated to the \emph{webSocket}. The server performs the same calculations in order to find a block that intersects the requested time, plays it and when finished the next part is automatically played without the user intervention.

	We observed differences in image quality when switching parts, that was even noticeable if we set a short block duration. We also noticed a small gap on audio when switching blocks but it was acceptable and speech recognition was not very affected.

	Back when we were implementing our solution, \ac{KMS} had not support for seeking videos, which meant that blocks would would always start playing from their beginning. This lead to a theoretic playing time error that can be at most half the duration of a block. In practice the maximum playing time error coincides with the block duration because we perform the intersection between the requested date and the block but we could decrease the error by half if we calculate the intersection with the requested time plus half the duration.

@@

	For playing video with an higher velocity we used \emph{ffmpeg}\footnote{\url{https://www.ffmpeg.org/}(accessed: 17 March 2016)} to convert the block into a new video with the desired velocity and seek time. Because the media duration is known, when the video started to being convert the headers located at the beginning of the file were already written and that made it possible to stream while converting.

	%{\color{red} [TALK ABOUT COMPLEXITY OF DECODE VS DECODE+MANIPULATE+ENCODE]}

	Although we implemented a solution that worked, we immediately noticed that \emph{ffmpeg} would take some time to initialize and that lead to pauses between switching parts.

	Later the \emph{Kurento} team released a version with support for seeking videos but we had to suspend the implementation of the fast forwarding feature as currently \ac{KMS} is not supporting that.

	We could implement fast forwarding without real time conversion by creating multiple versions of the same video with different velocities after the recording of a block. When a user needed to play he would also need to specify one of the available velocities. We did not followed this approach has it would require a bigger disk space usage.

@@

	One of our concerns during the development of our solution was the storage scalability. Saving files directly into file system would require an extra effort to distribute and replicate files among servers. For that purpose \emph{Kurento} team developed \emph{Kurento Repository}\footnote{\url{http://doc-kurento-repository.readthedocs.org} (accessed on 17 March 2016)} which is based on \emph{MongoDB}.

	One of the features that \emph{Kurento Repository} provides is the ability to play directly from the database without having to download the entire file to \ac{KMS}. The same is true for recording, but because the file headers are in the beginning and the file is written until is stops, the headers don't contain the necessary information for seeking the file.

	Although we gain with scalability with this approach, we lose access over the file for changing it to fit our needs, namely for using \emph{ffmpeg} or other video manipulation tool.

	Although we did not implemented recorded file seeking, that could be achieved by waiting for full file recording and then proceed to database insertion with the correct headers. Another approach would be the specification of the file duration before recording so the correct file headers could be written a priori. Both approaches were not possible to implement using just the \emph{Kurento} clients, we would need change the source code of \emph{Kurento} in order to add those new features.

>>>>

\section{Stream Recording}
In this section we describe our approaches for the stream recording implementation, namely recording on the client side and server side.
%RP Innitially we experimented with local recording. Even thought it got it to work, it consumed to much bandwith. We then tried ....  In this section, we describe the several techniques evaluated.

\subsection{Client side recording}
In order to record used shared streams, our first approach consisted in locally recording video and audio streams into blobs of limited duration. Because each user was recording directly from their shared local resources, we achieved the best stream quality.
%RP used -> user?

After recording, each block was uploaded to our server through \ac{HTTP}, saved into the server's file system, the block meta-data was created and inserted on our database. Each block meta-data contained the file's location for the recorded block, the starting date, duration, user identification and group identification.
%RP referenciar o modelo na secção 3.xx?
When the file was completely saved, the meta-data was advertised to the remaining users. This meta-data was simply used to refresh the user interface and was completely discarded after that, because an huge amount of small blocks' meta-data would use more and more memory as time went by.

	The process to play a video was quite simple. The user specified which user and date he intended to play, the server calculated the intersection between the requested date and the block bounds and returned the file to the client.
	Although this idea was fairly simple, we couldn't achieve seamless sequential block switching. Downloading the video file always took a noticeable amount of time. To solve this problem, while we were playing a block, the next one was downloading in parallel, so when the current block finished playing we would have an available block to play.

        With this technique, block switching became more acceptable, but switching the \ac{URL} always produced a flash. We solved this problem by having two layers and changing the \ac{URL} on the back layer. When the front layer froze on the last frame, we changed the back layer to the front.

	After we implemented this recording solution using this approach, we tested locally (client and server on the same machine) and remotely. For remote connections we observed a fairly high bandwidth usage, mainly because blocks were both sent and received at maximum quality. 
        %RP We deemed this high bandwidth usage to be a significant bottleneck on our application and decided to explore other solutions.

\subsection{Server side recording to file system}
%RP começar por explicar razão para usar KMS: permite gravar e juntar várias streams numa só, o que torna a aplicação muito mais escalável.

Before recording any type of stream, we had to analyze the user media offer in order to check if a video was really being received by \ac{KMS}, otherwise if we did not verify the user's offer, the recording video would be black.
%RP Entras logo em grandes pormenores. Falta aqui uma introdução.

The streaming content received on \ac{KMS} was already compressed due to \ac{WebRTC}'s exchanged quality of service metrics data. As a direct consequence, our recorder solution used on most cases less disk space per block, but would never use more storage than client side recording. \ac{KMS} allows recording files using \emph{webm} and \emph{mp4} containers.
%RP escolheste qual?

	With server side recording, the user would maintain always the same stream \ac{URL} even if it is playing real-time video or reproducing recorded video. It is \ac{KMS} that sends different content through that stream. When a user desires to play recorded video, a \emph{webSocket} message is sent specifying the time and the intended user id. The group identification is not sent because it is already associated to the \emph{webSocket}. The server performs the same calculations in order to find a block that intersects the requested time, plays it and when finished, the next part is automatically played without the user intervention.

	We still observed differences in image quality when switching parts. That was even noticeable if we set a short block duration. We also noticed a small gap on audio when switching blocks but it was acceptable and speech understanding was not very affected.

	When we started implementing our solution, \ac{KMS} had no support for seeking videos, which meant that blocks would would always start playing from their beginning. This lead to a theoretic playing time error that can be at most half the duration of a block. In practice, the maximum playing time error coincides with the block duration because we perform the intersection between the requested date and the block. But we could decrease the error by half if we calculate the intersection with the requested time plus half the duration.
        %RP já tinhamos falado sobre isto. O erro é no máximo o tamanho do bloco pois queres garantir que a pessoa ve sempre o que solicita. Assim começas sempre no bloco que fica antes e nunca no que fica depois.

        %RP Figura tem de ser apresentada no texto e explicada. O que significam as letras?

@@

        \ac{KMS} does not support playing video at higher speeds, a useful feature that enables users to watch a video in less time than its duration.
	For playing video with an higher velocity, we used \emph{ffmpeg}\footnote{\url{https://www.ffmpeg.org/}(accessed: 17 March 2016)} to convert the block into a new video with the desired velocity and seek time. Because the media duration is known, when the video started to being converted the headers located at the beginning of the file were already written and that made it possible to stream while converting.
	%{\color{red} [TALK ABOUT COMPLEXITY OF DECODE VS DECODE+MANIPULATE+ENCODE]}
	Although we implemented a solution that worked, we immediately noticed that \emph{ffmpeg} would take some time to initialize and that lead to pauses between switching parts.

	Later the \emph{Kurento} team released a version with support for seeking videos but we had to suspend the implementation of the fast forwarding feature as currently \ac{KMS} is not supporting that.
        %RP fiquei sem perceber. O método com o ffmpeg deixou de funcionar? Ou desististe devido ao outros problemas (delay, qualidade, cpu)?

	We could have implemented fast forwarding without real time conversion by creating multiple versions of the same video with different playback speeds after the recording of a block. When a user needed to play, he would also need to specify one of the available speed. We did not followed this approach has it would require larger disk space usage.

@@

	One of our concerns during the development of our solution was the storage scalability. Saving files directly into the file system would require an extra effort to distribute and replicate files among servers. For that purpose, the \emph{Kurento} team developed \emph{Kurento Repository}\footnote{\url{http://doc-kurento-repository.readthedocs.org} (accessed on 17 March 2016)} which is based on \emph{MongoDB}.

	One of the features that \emph{Kurento Repository} provides is the ability to play directly from the database without having to download the entire file to \ac{KMS}. The same is true for recording, but because the file headers are in the beginning and the file is written until is stops, the headers don't contain the necessary information for seeking the file.
	Although we gain with scalability with this approach, we lose access over the file for changing it to fit our needs, namely for using \emph{ffmpeg} or other video manipulation tool.

	Although we did not implement recorded file seeking, that could be achieved by waiting for full file recording and then proceed to database re-insertion with the correct headers. Another approach would be the specification of the file duration before recording, so the correct file headers could be written a priori. Both approaches were not possible to implement using just the \emph{Kurento} clients. We would need change the source code of \emph{Kurento} in order to add those new features.
        %RP explicar razões para não o fazer: perder upgrades, tornar instalação mais simples, maior complexidade manutenção app, etc.
